---
title: "Pwani University Boot Camp"
author: "Author : Moses Kioko"
date: 'Date : `r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_document:
    theme: united
    highlight: tango
    fontsize: 4pt
    code_folding: show
    toc: true
    number_sections: true
    code_download: true
    toc_float:
      collapsed: true
---

# Work space set up 
Prior to scripting in R, we removing the objects in R memory to pave way for a new objects, we do this by help of `rm()` function.

```{r}
#clear objects 
rm(list = ls(all=T))
```

For easier access of files in directory and organized saving objects in R memory we need to change default computer directory to a folder that we can easily trace and access. We do this my help of `setwd()` function.


```{r message=FALSE, warning=FALSE}
#setting working directory 

#set working directory 
setwd('D:/pwani data collection/eric/Revised Bootcamp/MAWW/Workshop')

# print the current working directory 
print(getwd())
```

# Packages

Installing packages can be done by downloading the packages from CRAN repository to R environment.*Please note; if you had prior installation of this packages you can skip re-installing them*

## Installing Packages 
+ Installing packages from CRAN 
  - install single package *i.e* `install.package("lubridate")`
  - install multiple packages by use of concatenator **c** *i,e* `install.packages(c('tidyverse','lubridate','forecast'))`
+ Installing Packages from Github 
  - install devtools package *i.e* `install.packages("devtools")`
  - load the library *i.e* `library(devtools)`
  - install the package *i.e* `devtools::install_github("tidyverse/lubridate")`

```{r message=FALSE, warning=FALSE}
#installing packages 
#install.packages(c('tidyverse','lubridate','janitor','viridis','forecast'), dependencies = TRUE)

```

Attached is a brief description of the packages that will be used in this session 

Package                   | Description 
--------------------------|------------------------------------------------------------------------------
Tidyverse Package         | Manipulating data and plotting data [link](https://tidyverse.tidyverse.org/)
Lubridate package         | Manipulating and changing date format [link](https://www.rdocumentation.org/packages/lubridate)
Janitor Package           | Cleaning column names [link](https://www.rdocumentation.org/packages/janitor/versions/2.0.1)
Viridis Package           | R color scale [link](https://www.rdocumentation.org/packages/viridis/versions/0.5.1)
Forecast Package          | Handy in time-series [link](https://www.rdocumentation.org/packages/forecast/versions/8.13/topics/Arima)



## Loading Packages 
To use the functions in a package you should load the package into R memory and you can do that by ;

+ loading single package *i.e* `library()` command
+ Load multiple packages *i.e*  [pacman package](https://www.rdocumentation.org/packages/pacman/versions/0.5.1)

```{r echo=TRUE, message=FALSE, warning=FALSE}
#load libraries
library(tidyverse) # manipulating data, plotting package 
library(lubridate) # changing date formats 
library(janitor) # cleaning variable names 
library(viridis) # coloring package 
library(forecast) # for time series 
```

# Importing Data 
Reading data into R memory using [readr package](https://www.rdocumentation.org/packages/readr/versions/1.3.1) housed in tidyverse package 

```{r echo=TRUE, message=FALSE, warning=FALSE}
#import temperature data
temp <- read_delim('temperature.csv',';',trim_ws = TRUE) %>% 
  clean_names()

#import precipitation data 
precipitation <- read_delim('precipitation.csv', ';',trim_ws = TRUE) %>% 
  clean_names()

#import MAWM data 
yield <- read_delim('MAWW_data.csv',';',trim_ws = TRUE) %>% 
  clean_names()
```

# Temperature Data
The are several steps that are involved to clean the temperature data ;

+ Change the date format using lubricate package 
+ Calculate the year from date 
+ Summarize the mean temperature per year and site 

```{r echo=TRUE, message=FALSE, warning=FALSE}
#summarizing data 
temp_per_site <- temp %>% 
  mutate(date = as_date(exceldate,origin ="1899-01-01")) %>% 
  mutate(year = year(date)) %>% 
  group_by(site,year) %>% 
  summarise(mean_temp = mean(value)) %>% 
    dplyr::filter(year<2006)
```

## Scatter Graph (loess fit)
To under understand general distribution of mean annual temperature we do a scatter plot with polynomial fit fuction 



```{r echo=TRUE, message=FALSE, warning=FALSE,fig.height=6,fig.width=9,fig.align='center'}
#Single plot the data 
loess_temp <- temp_per_site %>% 
  ggplot(aes(year,mean_temp))+
  geom_point(color ='#8A0707')+
  geom_smooth(stat = 'smooth',method = 'loess',color ='#c71585')+
  theme_minimal()+
  labs(title = "Mean Temperature",
       subtitle = "Distribution of Mean Annual Temperature (1950 ~ 2005)",
       y="Mean Temperature(degrees)",
       x="Year",
       caption = 'Author: Moses Kioko \n Profession: Data Analyst \n Pwani University Bootcamp @2020')+
  theme(legend.position = "bottom",
        legend.background = element_rect(color = "grey"),
        legend.title =  element_blank(),
        legend.text = element_text(family = "serif"),
        plot.title = element_text(hjust = 0.5,family = "serif"),
        axis.text = element_text(family = "serif", face = 'bold'),
        axis.title = element_text(family = "serif"),
        axis.line = element_line(colour = "grey30",size = .2),
        axis.ticks = element_line(colour = "grey30"),
        plot.subtitle = element_text(family = "serif",face = "italic",hjust = 0.5),
        plot.caption = element_text(size = 8,color = "grey",family = "serif",face = 'italic'))


#save the plot 
ggsave(plot= loess_temp,
       height = 8.1,
       width = 12,
       dpi = 1000,
       device = 'jpg',
       filename = 'Mean Annual Temperature Distribution Trend.jpg')

#print the plot 
print(loess_temp)
```

## Scatter Graph (linear fit)
To under understand trend of mean annual temperature distribution we do a scatter plot with linear fit function 

```{r echo=TRUE, message=FALSE, warning=FALSE,fig.height=6,fig.width=9,fig.align='center'}
#Single plot the data 
# please exclude the anything above 2006
lm_temp <- temp_per_site %>% 
  ggplot(aes(year,mean_temp))+
  geom_point(color ='#8A0707')+
  geom_smooth(stat = 'smooth',method = 'lm',color ='#c71585')+
  theme_minimal()+
  labs(title = "Mean Temperature",
       subtitle = "Distribution of Mean Annual Temperature (1950 ~ 2005)",
       y="Mean Temperature(degrees)",
       x="Year",
       caption = 'Author: Moses Kioko \n Profession: Data Analyst \n Pwani University Bootcamp @2020')+
  theme(legend.position = "bottom",
        legend.background = element_rect(color = "grey"),
        legend.title =  element_blank(),
        legend.text = element_text(family = "serif"),
        plot.title = element_text(hjust = 0.5,family = "serif"),
        axis.text = element_text(family = "serif", face = 'bold'),
        axis.title = element_text(family = "serif"),
        axis.line = element_line(colour = "grey30",size = .2),
        axis.ticks = element_line(colour = "grey30"),
        plot.subtitle = element_text(family = "serif",face = "italic",hjust = 0.5),
        plot.caption = element_text(size = 8,color = "grey",family = "serif",face = 'italic'))


#save the plot 
ggsave(plot= lm_temp,
       height = 8.1,
       width = 12,
       dpi = 1000,
       device = 'jpg',
       filename = 'Mean Annual Temperature Distribution Linear Regression plot.jpg')

#print the plot 
print(lm_temp)
```

## Box Plot 

```{r echo=TRUE, message=FALSE, warning=FALSE,fig.height=7,fig.width=9,fig.align='center'}
#distribution of temperature per site
box_temp <- temp_per_site %>% 
  ggplot(aes(site,mean_temp,color =site))+
  geom_boxplot(outlier.shape = 18,
               outlier.colour = "red",
               outlier.alpha = 1,
               outlier.size = 2.5)+
  geom_jitter(shape=16, position=position_jitter(0.2),cex=0.9)+
  scale_color_viridis_d()+
  theme_minimal()+
  labs(title = "Mean Temperature",
       subtitle = "Distribution of Mean Annual Temperature per Site (1950 - 2005)",
       y="Mean Temperature (degrees)",
       x="Agricultural Site",
       caption = 'Author: Moses Kioko \n Profession: Data Analyst \n Pwani University Bootcamp @2020')+
  theme(legend.position = "none",
        legend.background = element_rect(color = "grey"),
        legend.title =  element_blank(),
        legend.text = element_text(family = "serif"),
        plot.title = element_text(hjust = 0.5,family = "serif"),
        axis.text = element_text(family = "serif", face = 'italic'),
        axis.title = element_text(family = "serif"),
        plot.subtitle = element_text(family = "serif",face = "italic",hjust = 0.5),
        plot.caption = element_text(size = 8,color = "grey",family = "serif",face = 'italic'))



#save the plot 
ggsave(plot= box_temp,
       height = 8.1,
       width = 12,
       dpi = 1000,
       device = 'jpg',
       filename = 'Mean Annual Temperature Distribution boxplot plot.jpg')


#plot the graph 
print(box_temp)

```

## Bar Graph 

```{r echo=TRUE, message=FALSE, warning=FALSE,fig.height=7,fig.width=9,fig.align='center'}
#distribution of temperature per site
temp_bar_graph <- temp_per_site %>% 
  ggplot(aes(year,mean_temp,fill =site))+
  geom_col(position = 'dodge')+
  facet_wrap(~site)+
  scale_fill_viridis_d()+
  theme_bw()+
  labs(title = "Mean Temperature",
       subtitle = "Distribution of Mean Annual Temperature per Site (1950 - 2005)",
       y="Mean Temperature (degrees)",
       x="Agricultural Site",
       caption = 'Author: Moses Kioko \n Profession: Data Analyst \n Pwani University Bootcamp @2020')+
  theme(legend.position = "none",
        legend.background = element_rect(color = "grey"),
        legend.title =  element_blank(),
        legend.text = element_text(family = "serif"),
        plot.title = element_text(hjust = 0.5,family = "serif"),
        axis.text = element_text(family = "serif", face = 'italic'),
        axis.title = element_text(family = "serif"),
        plot.subtitle = element_text(family = "serif",face = "italic",hjust = 0.5),
        plot.caption = element_text(size = 8,color = "grey",family = "serif",face = 'italic'))



#save the plot 
ggsave(plot= temp_bar_graph,
       height = 8.1,
       width = 12,
       dpi = 1000,
       device = 'jpg',
       filename = 'Mean Annual Temperature Distribution bar graph.jpg')

#plot the graph 
print(temp_bar_graph)

```


# Precipitation Data 
The following steps are used to tidy the precipitation data;

+ Change the date format using lubricate package 
+ Calculate the year from date 
+ Summarize the mean precipitation per year and site 

```{r echo=TRUE, message=FALSE, warning=FALSE}
#cleaning precipitation data by formatting date and calculating mean precipitation per year and site 
precip_per_site <- precipitation %>% 
  mutate(date = as_date(exceldate,origin ="1899-01-01")) %>% 
  mutate(year = year(date)) %>% 
  group_by(site,year) %>% 
  summarise(mean_precip = mean(value)) %>% 
  clean_names() %>% 
  dplyr::filter(year<2006)
```

## Scatter Plot (loess fit)

```{r echo=TRUE, message=FALSE, warning=FALSE,fig.height=7,fig.width=9,fig.align='center'}
#Single plot the data 
prep_polynominal <- precip_per_site %>% 
  ggplot(aes(year,mean_precip))+
  geom_point(color ='blue')+
  geom_smooth(stat = 'smooth',method = 'loess',color ='steelblue')+
  theme_minimal()+
  labs(title = "Mean Precipitation",
       subtitle = "Distribution of Mean Annual Precipitation (1950 ~ 2005)",
       y="Mean Precipitation",
       x="Year",
       caption = 'Author: Moses Kioko \n Profession: Data Analyst \n Pwani University Bootcamp @2020')+
  theme(legend.position = "bottom",
        legend.background = element_rect(color = "grey"),
        legend.title =  element_blank(),
        legend.text = element_text(family = "serif"),
        plot.title = element_text(hjust = 0.5,family = "serif"),
        axis.text = element_text(family = "serif", face = 'italic'),
        axis.title = element_text(family = "serif"),
        axis.line = element_line(colour = "grey30",size = .2),
        axis.ticks = element_line(colour = "grey30"),
        plot.subtitle = element_text(family = "serif",face = "italic",hjust = 0.5),
        plot.caption = element_text(size = 8,color = "grey",family = "serif",face = 'italic'))



#save the plot 
ggsave(plot= prep_polynominal,
       height = 8.1,
       width = 12,
       dpi = 1000,
       device = 'jpg',
       filename = 'Mean Annual Precipitation Distribution polynomial fit graph.jpg')

#plot the graph 
print(prep_polynominal)
```


## Scatter plot (Linear fit)

```{r echo=TRUE, message=FALSE, warning=FALSE,fig.height=7,fig.width=9,fig.align='center'}
#Single plot the data 
prep_linear <- precip_per_site %>% 
  ggplot(aes(year,mean_precip))+
  geom_point(color ='blue')+
  geom_smooth(stat = 'smooth',method = 'lm',color ='steelblue')+
  theme_minimal()+
  labs(title = "Mean Precipitation",
       subtitle = "Distribution of Mean Annual Precipitation (1950 ~ 2005)",
       y="Mean Precipitation",
       x="Year",
       caption = 'Author: Moses Kioko \n Profession: Data Analyst \n Pwani University Bootcamp @2020')+
  theme(legend.position = "bottom",
        legend.background = element_rect(color = "grey"),
        legend.title =  element_blank(),
        legend.text = element_text(family = "serif"),
        plot.title = element_text(hjust = 0.5,family = "serif"),
        axis.text = element_text(family = "serif", face = 'italic'),
        axis.title = element_text(family = "serif"),
        axis.line = element_line(colour = "grey30",size = .2),
        axis.ticks = element_line(colour = "grey30"),
        plot.subtitle = element_text(family = "serif",face = "italic",hjust = 0.5),
        plot.caption = element_text(size = 8,color = "grey",family = "serif",face = 'italic'))



#save the plot 
ggsave(plot= prep_linear,
       height = 8.1,
       width = 12,
       dpi = 1000,
       device = 'jpg',
       filename = 'Mean Annual Precipitation Distribution Linear fit graph.jpg')

#plot the graph 
print(prep_linear)
```

## Box Plot


```{r echo=TRUE, message=FALSE, warning=FALSE,fig.height=7,fig.width=9,fig.align='center'}
#distribution of precipitation per site
scatter_boxplot_prep <- precip_per_site %>% 
  ggplot(aes(site,mean_precip,color =site))+
  geom_boxplot(outlier.shape = 18,
               outlier.colour = "red",
               outlier.alpha = 1,
               outlier.size = 2.5)+
  geom_jitter(shape=16, position=position_jitter(0.2),cex=0.9)+
  scale_color_viridis_d()+
  theme_minimal()+
  labs(title = "Mean Precipitation",
       subtitle = "Distribution of Mean Annual Precipitation per Site (1950 - 2005)",
       y="Mean Precipitation",
       x="Agricultural Site",
       caption = 'Author: Moses Kioko \n Profession: Data Analyst \n Pwani University Bootcamp @2020')+
  theme(legend.position = "none",
        legend.background = element_rect(color = "grey"),
        legend.title =  element_blank(),
        legend.text = element_text(family = "serif"),
        plot.title = element_text(hjust = 0.5,family = "serif"),
        axis.text = element_text(family = "serif", face = 'italic'),
        axis.title = element_text(family = "serif"),
        plot.subtitle = element_text(family = "serif",face = "italic",hjust = 0.5),
        plot.caption = element_text(size = 8,color = "grey",family = "serif",face = 'italic'))


#save the plot 
ggsave(plot= scatter_boxplot_prep,
       height = 8.1,
       width = 12,
       dpi = 1000,
       device = 'jpg',
       filename = 'Mean Annual Precipitation Distribution box plot.jpg')

#plot the graph 
print(scatter_boxplot_prep)
```

## Bar Graph

```{r echo=TRUE, message=FALSE, warning=FALSE,fig.height=7,fig.width=9,fig.align='center'}
#distribution of precipitation per site
prep_bar <- precip_per_site %>% 
  ggplot(aes(year,mean_precip,fill =site))+
  geom_col(position = 'dodge')+
  facet_wrap(~site)+
  scale_fill_viridis_d()+
  theme_bw()+
  labs(title = "Mean Precipitation",
       subtitle = "Distribution of Mean Annual Precipitation per Site (1950 - 2005)",
       y="Mean Precipitation (mm)",
       x="Agricultural Site",
       caption = 'Author: Moses Kioko \n Profession: Data Analyst \n Pwani University Bootcamp @2020')+
  theme(legend.position = "none",
        legend.background = element_rect(color = "grey"),
        legend.title =  element_blank(),
        legend.text = element_text(family = "serif"),
        plot.title = element_text(hjust = 0.5,family = "serif"),
        axis.text = element_text(family = "serif", face = 'italic'),
        axis.title = element_text(family = "serif"),
        plot.subtitle = element_text(family = "serif",face = "italic",hjust = 0.5),
        plot.caption = element_text(size = 8,color = "grey",family = "serif",face = 'italic'))

#save the plot 
ggsave(plot= prep_bar,
       height = 8.1,
       width = 12,
       dpi = 1000,
       device = 'jpg',
       filename = 'Mean Annual Precipitation bar graph.jpg')

#plot the graph 
print(prep_bar)
```

# Modelling Germany Wheat Data
Linear models help us to generation a mathematical equation to map a relationship between two or more variables. We are going to use linear model to model the wheat data, there fore a couple of packages are going to be helpful in this section 

Package                   | Description
--------------------------|------------------------------------------------------
GGally package            | Plotting correlation table [link](https://www.rdocumentation.org/packages/GGally/versions/1.5.0)
MASS Package              | helps selecting most important variables for the model [link](https://www.rdocumentation.org/packages/MASS/versions/7.3-53) 

```{r echo=TRUE, message=FALSE, warning=FALSE}
#install packages 
#install.packages('MASS',dependencies = T)
#install.packages('GGally',dependencies = T)

#load the packages 
library(MASS)
library(GGally)
```


This section we;

  + Select the data we would want to include in our model
  + Rename the variables

```{r echo=TRUE, message=FALSE, warning=FALSE}
#rename variables 
model_data <- yield %>% 
  dplyr::select(yield = y_max,
                altitude = alt,
                fertilizer = n_fert,
                `soil quality` = sq,
                precipitation =map,
                temperature = mat,acwb) 

#save model names 
model_names <- names(model_data[,2:6])
```

Plotting correlation plot

+ Why correlation plot 
  - Distribution of data points within a variable 
  - Get the correlation coefficients that informs on whether to drop or maintain the variable 
    + Helps avoid Auto correlation problem 
    + Feature engineering *(i.e advanced topic of excluding highly related variables as they don not inform much on prediction)*

```{r echo=TRUE, message=FALSE, warning=FALSE,fig.align='center',fig.height=6,fig.height=6}
#correlation plot 
cor_plot <- GGally::ggpairs(model_data)

#save the plot 
ggsave(plot= cor_plot,
       height = 8.1,
       width = 12,
       dpi = 1000,
       device = 'jpg',
       filename = 'Correlation Plot.jpg')

#plot the graph 
print(cor_plot)
```

Linear Regression Assumption

  + Normality of the y  variable ..etc

To ensure we satisfy this requirement we do a density plot for wheat yield. 
please note ; 

  + Blue curve signifies the normal expected curve 
  + Grey curve indicates the actual distribution of wheat yield 


```{r echo=TRUE, message=FALSE, warning=FALSE, fig.align='center',fig.height=6,fig.width=9}
density_plot <- model_data %>% 
  ggplot(aes(yield))+
  geom_density(size =.8,color ='grey30',fill ='grey')+
  stat_function(fun=dnorm,color="blue",size =.8,args=list(mean=mean(model_data$yield), 
                                  sd=sd(model_data$yield)))+
  theme_minimal()+
  labs(title = "Yield Distribution",
       subtitle = "Normality check for wheat Distribution (1958 - 2009)",
       y="",
       x="",
       caption = 'Author: Moses Kioko \n Profession: Data Analyst \n Pwani University Bootcamp @2020')+
  theme(legend.position = "none",
        legend.background = element_rect(color = "grey"),
        legend.title =  element_blank(),
        legend.text = element_text(family = "serif"),
        plot.title = element_text(hjust = 0.5,family = "serif"),
        axis.text = element_text(family = "serif", face = 'italic'),
        axis.title = element_text(family = "serif"),
        plot.subtitle = element_text(family = "serif",face = "italic",hjust = 0.5),
        plot.caption = element_text(size = 8,color = "grey",family = "serif",face = 'italic'))

#save the plot 
ggsave(plot= density_plot,
       height = 8.1,
       width = 12,
       dpi = 1000,
       device = 'jpg',
       filename = 'Wheat Yield Distribution Graph.jpg')

#plot the graph 
print(density_plot)
```


Running Linear/Multiple regression we use base R statistical function `lm()`. We run a regression model between the following variables **`r model_names`**. Please note; We regress the **yield** variable against the rest of the variables to establish the relationship.

```{r echo=TRUE, message=FALSE, warning=FALSE,fig.align='center',fig.width=9}
#linear model 
model = lm(yield ~ altitude+fertilizer+`soil quality`+precipitation+temperature,data=model_data) 


#print out the summary of the model
summary(model)
```

To determine the best model we deploy forward backward or both methods to determine the variables that highly contribute to the model, this approach uses variance to determine the variables.

```{r echo=TRUE, message=FALSE, warning=FALSE,fig.align='center',fig.width=9}
#compute the best model
best_model = MASS::stepAIC(model,direction = "both")

#print out the summary of the model
summary(best_model)

```



The final model equation $Yield = 41.049 + 0.167Ferterlizer + 0.243SoilQuality + 0.019Precipitation - 3.603Temperature$
